---
title: "Minimum Detectable Effect (MDE) Calculation"
author: "Aleksei Prishchepo"
date: "2025-08-07"
categories: ["A/B Testing", Product, Statistics, R]
image: "image.svg"
fig-format: svg
# execute:
#   echo: false
---

Minimum Detectable Effect (MDE) is defined as the smallest difference
between a control and a test group that your A/B test can reliably
identify as statistically significant. It's a critical concept because
it helps in determining the necessary sample size for an experiment and
in interpreting the results.

## How MDE is calculated in experimental design

The MDE calculation depends on several key parameters:

-   Sample size ($n$ for the test group, $m$ for the control group).
-   Significance level (alpha, $\alpha$): This is the probability of
    a Type I error (falsely rejecting the null hypothesis), typically
    set at 0.05.
-   Statistical power (1-beta, $1-\beta$): This is the probability of
    correctly detecting an effect when one truly exists, commonly set at
    0.8 (or 80%).
-   Variance ($\sigma^2$) of the metric being measured in the
    population, often estimated from historical data.
-   Ratio of control to test group sizes ($k = m/n$).

```{r plot-normal-distribution, echo=FALSE, fig.cap="Control and test distributions with critical values, alpha, beta, and power shaded."}
#| message: false
#| warning: false
#| fig-height: 4
#| fig-width: 7

library(ggplot2)
library(dplyr)
library(latex2exp)

controlDistribution <- data.frame(
  x = seq(-4, 7, by = 0.01),
  y = dnorm(seq(-4, 7, by = 0.01))
)
testDistribution <- data.frame(
  x = seq(-4, 7, by = 0.01),
  y = dnorm(seq(-4, 7, by = 0.01), mean = 3) # Example shift for test group
)
criticalValues <- qnorm(c(.025, .975))

shadeAlpha <- rbind(
  c(criticalValues[1], 0),
  subset(controlDistribution, x < criticalValues[1])
)
shadeBeta <- rbind(
  c(criticalValues[2], 0),
  subset(testDistribution, x < criticalValues[2])
)
shadePower <- rbind(
  c(criticalValues[2], 0),
  subset(testDistribution, x > criticalValues[2])
)

ggplot(controlDistribution, aes(x, y)) +
  geom_polygon(data = shadeAlpha, aes(x = x, y = y, fill = "alpha / 2", alpha = 0.5)) +
  geom_polygon(data = shadeBeta, aes(x = x, y = y, fill = "beta", alpha = 0.5)) +
  geom_polygon(data = shadePower, aes(x = x, y = y, fill = "power", alpha = 0.5)) +
  geom_line(
    color = "blue",
    linewidth = 1
  ) +
  geom_line(
    data = testDistribution,
    aes(x, y),
    color = "red",
    linewidth = 1
  ) +
  geom_vline(xintercept = criticalValues[1], linetype = "dashed", linewidth = 0.75) +
  geom_vline(xintercept = criticalValues[2], linetype = "dashed", linewidth = 0.75) +
  labs(
    title = TeX("$\\alpha=0.05, \\beta=0.2, {power}=0.8$"),
    x = "",
    y = "Density",
    fill = ""
  ) +
  theme_classic() +
  theme(panel.border = element_rect(colour = "black", fill = NA, linewidth = 1.5)) +
  guides(alpha = "none")
```

## Calculating MDE for a given sample size

If you have a fixed number of users (n) or a defined sample size (m or
n), you can determine the smallest effect (e) that your test can
reliably detect:

$$e > \sqrt{\dfrac{(z_{1-\alpha/2} + z_{1-\beta})^2 (1 + k)\sigma^2}{m}}$$

$z_{1-\alpha/2}$ and $z_{1-\beta}$ are the Z-scores corresponding to the
desired significance level and power, respectively.

$k=\frac{m}{n}$ is the ratio of the control group size to the test group
size (e.g., $k=1$ for a 1:1 split).

$\sigma^2$ is the estimated variance of the metric.

$e$ is the minimal effect.

### How to calculate MDE in R

For instance, with 100,000 total users, a sigma of 500, a k-ratio of 2,
alpha = 0.05, and beta = 0.2, the minimum detectable effect would be
approximately 9.397 units, which translates to a 1.88% change from the
mean of 500.

```{r}
# Function to calculate MDE
calculate_mde <- function(n, alpha = 0.05, beta = 0.2, sigma = 1, k = 1) {
  z_alpha <- qnorm(1 - alpha / 2)
  z_beta <- qnorm(1 - beta)
  m <- n * k / (k + 1)
  e <- sqrt((z_alpha + z_beta)^2 * (1 + k) * sigma^2 / m)
  return(e)
}

calculate_mde(n = 100000, alpha = 0.05, beta = 0.2, sigma = 500, k = 2)
```

### Interactive MDE calculator

Below is an interactive calculator for finding the MDE based on your
sample size, alpha, beta, sigma, and k-ratio. Adjust the parameters to
see how they affect the MDE.

<iframe src="https://a13ks3i.shinyapps.io/mde-calculator/" width="100%" height="600" title="MDE Calculator">

</iframe>

## Finding the required sample size for a given MDE

The formula used to determine the necessary sample size (e.g., for the
control group, $m$) to detect a specific MDE ($e$) is:

$$m > \dfrac{(z_{1-\alpha/2} + z_{1-\beta})^2 (1 + k)\sigma^2}{e^2}$$

### How to calculate sample size in R

Example: if a monetary metric has a mean of 500 and a standard deviation
(sigma) of 500, and you want to detect a 2% effect (MDE = 10) with alpha
= 0.05 and beta = 0.2, and a test-to-control ratio of k = 2, the
required sample size would be approximately 29,434 users in the test
group and 58,867 users in the control group, totaling 88,301 users.

```{r}
# Function to calculate required sample size
calculate_sample_size <- function(e, alpha = 0.05, beta = 0.2, sigma = 1, k = 1) {
  z_alpha <- qnorm(1 - alpha / 2)
  z_beta <- qnorm(1 - beta)
  m <- (z_alpha + z_beta)^2 * (1 + k) * sigma^2 / e^2
  return(ceiling(m))
}
m <- calculate_sample_size(e = 10, alpha = 0.05, beta = 0.2, sigma = 500, k = 2)
n <- ceiling(m / 2)
total_users <- m + n

c(m, n, total_users)
```

### Interactive sample size calculator

Below is an interactive calculator that allows you to input your desired
MDE, alpha, beta, sigma, and k-ratio to compute the required sample
size. Adjust the parameters to see how they affect the sample size
needed for your A/B test.

<iframe src="https://a13ks3i.shinyapps.io/sample-size-calculator/" width="100%" height="600" title="Sample Size Calculator">

</iframe>

## Conclusion

A/B testing is a powerful tool for product development, and
understanding the concept of Minimal Detectable Effect (MDE) is crucial
for designing effective experiments. By calculating the required sample
size or MDE, you can ensure that your tests are statistically sound and
capable of providing meaningful insights into user behavior and product
performance.

```{r plot-image-svg, echo=FALSE}
#| message: false
#| warning: false

library(here)

controlDistribution <- data.frame(
  x = seq(-4, 7, by = 0.01),
  y = dnorm(seq(-4, 7, by = 0.01))
)
testDistribution <- data.frame(
  x = seq(-4, 7, by = 0.01),
  y = dnorm(seq(-4, 7, by = 0.01), mean = 3) # Example shift for test group
)
criticalValues <- qnorm(c(.025, .975))

shadeAlpha <- rbind(
  c(criticalValues[1], 0),
  subset(controlDistribution, x < criticalValues[1])
)
shadeBeta <- rbind(
  c(criticalValues[2], 0),
  subset(testDistribution, x < criticalValues[2])
)
shadePower <- rbind(
  c(criticalValues[2], 0),
  subset(testDistribution, x > criticalValues[2])
)

p <- ggplot(controlDistribution, aes(x, y)) +
  geom_polygon(data = shadeAlpha, aes(x = x, y = y, fill = "alpha / 2"), alpha = 0.5) +
  geom_polygon(data = shadeBeta, aes(x = x, y = y, fill = "beta"), alpha = 0.5) +
  geom_polygon(data = shadePower, aes(x = x, y = y, fill = "power"), alpha = 0.5) +
  geom_line(
    color = "blue",
    linewidth = 1
  ) +
  geom_line(
    data = testDistribution,
    aes(x, y),
    color = "red",
    linewidth = 1
  ) +
  geom_vline(xintercept = criticalValues[1], linetype = "dashed", linewidth = 0.75) +
  geom_vline(xintercept = criticalValues[2], linetype = "dashed", linewidth = 0.75) +
  theme_minimal() +
  theme(legend.position = "none", axis.title = element_blank())

ggsave(
  here(
    "posts", "20250807-mde", "image.svg"
  ),
  plot = p,
  units = "px",
  width = 1200,
  height = 1000,
  dpi = 300
)
```
