---
title: "Time-Series Clustering with R's dtwclust Package"
author: "Aleksei"
date: "2025-08-27"
categories: [Time-Series, Clustering, R]
image: "image.svg"
fig-format: svg
number-sections: true
---

This article is a practical guide for time-series clustering using the
`dtwclust` package. The `dtwclust` package in R (see
[vignette](https://cran.r-project.org/web/packages/dtwclust/vignettes/dtwclust.pdf){target="_blank"})
provides a powerful and flexible framework for time-series clustering,
allowing you to implement and compare various algorithms, particularly
those leveraging Dynamic Time Warping (DTW). This showcase will guide
you through a practical example of time-series clustering using
`dtwclust`, including data preparation, clustering execution,
visualization, and evaluation.

## What is Dynamic Time Warping (DTW)

[Dynamic Time Warping
(DTW)](https://en.wikipedia.org/wiki/Dynamic_time_warping){target="_blank"}
is a prominent distance measure used in shape-based time-series
clustering. Unlike Euclidean distance, which compares points at the same
time index, DTW allows for "warping" or stretching/compressing the time
axis of one series to find an optimal alignment with another. This
enables it to accurately measure similarity between time-series that may
vary in speed, length, or have phase shifts, but exhibit similar overall
shapes.

![Sample alignment performed by the DTW algorithm between two series.
The dashed blue lines exemplify how some points are mapped to each
other, which shows how they can be warped in time. Note that the
vertical position of each series was artificially altered for
visualization. Credits: Alexis
Sardá-Espinosa](dtw-alignment.svg){width="5in"}

## Data Preparation

```{r setup, include=FALSE, eval=FALSE}
# Install packages if not already installed
install.packages("bigmemory")
install.packages("dtwclust")
install.packages("httr2")
install.packages("tibble")
install.packages("doParallel")
install.packages("dplyr")
install.packages("tidyr")
```

We will use daily closing prices of various cryptocurrencies from the
dYdX exchange. The data will be fetched using the `httr2` package.

::: callout-note
**dYdX Exchange** is a decentralized finance (DeFi) platform that allows
users to trade perpetual derivatives, margin, and spot crypto assets
without a centralized intermediary.
:::

Let's start by fetching the list of available perpetual markets from the
dYdX API. We will then filter for popular cryptocurrencies and retrieve
their daily closing prices.

```{r fetch-markets}
#| message: false
#| warning: false
library(httr2)
library(tibble)
library(dplyr)
library(tidyr)

base_url <- "https://indexer.dydx.trade/v4"
full_url <- paste0(base_url, "/perpetualMarkets")

req <- request(full_url) |>
  req_headers(Accept = "application/json")

resp <- req_perform(req)
body <- resp_body_json(resp, simplifyVector = TRUE)

markets <- bind_rows(body$markets) |> as_tibble()
markets <- markets |> mutate(volume24H = as.numeric(volume24H))
markets <- markets |> arrange(desc(volume24H)) 

markets |> select(ticker, oraclePrice, priceChange24H, volume24H) |> head(5)
```

Let's take 30 the most traded markets.

```{r select-tickers}
tickers <- lapply(body$markets[1:31], function(market) {
  market$ticker
})
# remove MATIC-USD due to missing data
tickers <- tickers[!tickers %in% c("MATIC-USD")]
tickers <- tickers |>
  unlist() |>
  array()
```

Below are are helper functions to convert date-times to the required
format and to fetch daily candle data from the dYdX API. For some reason
at the time of writing this, the API ignores the `fromIso` parameter, so
we will filter the data after fetching it.

```{r}
to_iso8601_ns_utc <- function(datetime) {
  datetime_utc <- as.POSIXct(datetime, tz = "UTC")
  format(datetime_utc, format = "%Y-%m-%dT%H:%M:%OS9Z", usetz = FALSE)
}

get_dydx_daily_candles <- function(
    market,
    from_iso = NULL,
    to_iso = NULL,
    limit = NULL) {
  base_url <- "https://indexer.dydx.trade/v4"
  full_url <- paste0(base_url, "/candles/perpetualMarkets/", market)

  req <- request(full_url) |>
    req_headers(Accept = "application/json") |>
    req_url_query(resolution = "1DAY")

  from_iso <- if (!is.null(from_iso)) to_iso8601_ns_utc(from_iso) else NULL
  to_iso <- if (!is.null(to_iso)) to_iso8601_ns_utc(to_iso) else NULL

  if (!is.null(from_iso)) req <- req |> req_url_query(fromIso = from_iso)
  if (!is.null(to_iso)) req <- req |> req_url_query(toIso = to_iso)
  if (!is.null(limit)) req <- req |> req_url_query(limit = limit)

  resp <- req_perform(req)
  body <- resp_body_json(resp, simplifyVector = TRUE)

  df <- body$candles
  df <- df |> mutate(
    startedAt = as.POSIXct(startedAt,
      format = "%Y-%m-%dT%H:%M:%OSZ",
      tz = "UTC"
    ),
    open = as.numeric(open),
    high = as.numeric(high),
    low = as.numeric(low),
    close = as.numeric(close)
  )
  df |> as_tibble()
}

get_dydx_daily_candles("BTC-USD",
  from_iso = "2025-01-01T00:00:00Z",
  to_iso = "2025-08-26T00:00:00Z"
) |> head()
```

Now, we will fetch daily closing prices for our selection of popular
cryptocurrencies.

```{r fetch-crypto-data}
cryptos_list <- lapply(tickers, function(coin) {
  # cat("Fetching data for:", coin, "\n")
  df <- get_dydx_daily_candles(coin,
    from_iso = "2025-01-01T00:00:00Z",
    to_iso = "2025-08-26T00:00:00Z"
  )
  df <- df |> filter(startedAt >= as.POSIXct("2025-01-01", tz = "UTC"))
  df <- df |> select(ds = startedAt, y = close) |> arrange(ds)
  # z-score normalization
  df <- df |> mutate(y = (y - mean(y, na.rm = TRUE)) / sd(y, na.rm = TRUE))
  df$coin <- toupper(coin) %>% gsub("-USD", "", .)
  return(df)
})

cryptos_list <- bind_rows(cryptos_list)
```

Next, we will reshape the data into a wide format suitable for
clustering, where each row represents a cryptocurrency and each column
represents a daily closing price.

```{r reshape-data}
cryptos_list_wide <- cryptos_list  |> 
  pivot_wider(names_from = coin, values_from = y) |> 
  arrange(ds) |> 
  select(-ds) |>  
  as.list()
```

## Performing Hierarchical Clustering

We will perform hierarchical clustering using the DTW distance and the
"ward.D2" agglomeration method. Hierarchical clustering builds a
hierarchy of groups without requiring a pre-specified number of clusters
initially, and the process is deterministic.

-   `k = 4` specifies the desired number of clusters.

-   `type = "hierarchical"` sets the clustering algorithm type.

-   `distance = "dtw"` uses Dynamic Time Warping distance.

-   `seed = 42` for reproducibility of random initializations (if
    applicable).

-   `control = hierarchical_control(method = "ward.D2")` specifies the
    linkage method.

-   `args = tsclust_args(dist = list(window.size = 7))` sets DTW window
    constraint.

```{r hierarchical-clustering}
#| message: false
#| warning: false
library(dtwclust)

# Perform hierarchical clustering
hc_4_ward <- tsclust(cryptos_list_wide,
  k = 4,
  type = "hierarchical",
  distance = "dtw",
  seed = 42,
  control = hierarchical_control(method = "ward.D2"),
  args = tsclust_args(dist = list(window.size = 7))
)

# View the clustering summary
hc_4_ward
```

The output provides details about the clustering, including the distance
measure, centroid method, linkage method, and cluster sizes with their
average intra-cluster distances.

## Accessing Clustering Results

The `tsclust()` function returns an S4 object of class `TSClusters`. You
can access its slots, such as the cluster assignments, using the \@
operator.

```{r access-cluster-assignments}
# View cluster assignments for each time series
hc_4_ward@cluster
```

## Visualizing Clustering Results

The `plot()` method for `TSClusters` objects offers various
visualization types.

### Dendrogram

A dendrogram visually represents the hierarchy of clusters.

```{r fig-dendrogram-2}
#| fig-cap: "Dendrogram of hierarchical clustering using DTW distance and Ward.D2 linkage."
par(mar = c(0, 4, 2, 2)) # Adjust margins for better plot
plot(hc_4_ward,
  xlab = "", ylab = "", sub = "",
  main = "Hierarchical Clustering Dendrogram (DTW, Ward.D2)"
)
```

### Series and Centroids

Visualize the time series grouped by cluster, along with their
representative prototypes (centroids). By default, prototypes for
hierarchical clustering with PAM centroids are actual series from the
data.

```{r fig-series-centroids-2}
#| fig-cap: "Time series clustered with their centroids using DTW distance and Ward.D2 linkage."
plot(hc_4_ward, type = "sc") # sc = series + centroids
```

You can also plot a specific centroid, and even customize its
appearance.

```{r fig-specific-centroid}
#| fig-cap: Specific centroid (cluster 1).
#| fig-height: 3
#| fig-width: 4
plot(hc_4_ward,
  type = "centroids", clus = 1,
  linetype = 1, size = 1, alpha = 0.8
)
```

## Comparing Multiple Clustering Solutions and Evaluation

In practice, choosing the optimal number of clusters (`k`) and other
parameters is crucial. `dtwclust` allows you to test multiple
configurations simultaneously and evaluate them using Cluster Validity
Indices (CVIs).

::: callout-note
**Cluster Validity Indices** are quantitative metrics used to assess the
quality and "purity" of clustering results. Since clustering is often an
unsupervised process, CVIs provide an objective way to evaluate
performance, especially when comparing different clustering algorithms
or configurations.
:::

To accelerate the process, especially when testing many combinations,
parallelization is highly recommended. `dtwclust` integrates with the
`foreach` and `doParallel` packages for this purpose.

```{r parallel-clustering-evaluation}
#| message: false
#| warning: false
library(bigmemory)
library(doParallel)

# Define a range of k values and agglomeration methods to test
k_values <- 3:6
linkage_methods <- c("ward.D2", "average", "single", "complete")

# Initialize a parallel backend
# Use detectCores() - 1 to leave one core free
num_cores <- detectCores() - 1
if (num_cores < 1) num_cores <- 1 # Ensure at least one core is used

cl <- makeCluster(num_cores)
registerDoParallel(cl)

# Perform multiple hierarchical clusterings in parallel
hc_par <- tsclust(cryptos_list_wide,
  k = k_values,
  type = "hierarchical",
  distance = "dtw",
  seed = 42,
  control = hierarchical_control(method = linkage_methods),
  args = tsclust_args(dist = list(window.size = 7)),
  trace = TRUE
)

# Stop the parallel cluster and revert to sequential computation
stopCluster(cl)
registerDoSEQ()
```

### Evaluate the results using internal CVIs

We'll use [Silhouette
(Sil)](https://en.wikipedia.org/wiki/Silhouette_(clustering)){target="_blank"},
[Dunn (D)](https://en.wikipedia.org/wiki/Dunn_index){target="_blank"},
and [Calinski-Harabasz
(CH)](https://en.wikipedia.org/wiki/Calinski%E2%80%93Harabasz_index){target="_blank"}
indices. Higher values generally indicate better clustering for these
indices.

```{r evaluate-clusterings}
cvi_results <- lapply(hc_par, cvi, type = c("Sil", "D", "CH")) %>%
  do.call(rbind, .)

# Find the configuration that maximizes each CVI
optimal_indices <- apply(cvi_results, MARGIN = 2, FUN = which.max)
```

#### Display CVI results and optimal configurations

```{r display-cvi-results}
print(cvi_results)
```

```{r display-optimal-configurations}
print(optimal_indices)
```

#### Retrieve the best clustering based on Silhouette index

Let's extract the clustering that achieved the highest Silhouette score.

```{r best-clustering-silhouette}
best_clustering_sil <- hc_par[[optimal_indices["Sil"]]]
best_clustering_sil
```

#### Retrieve the best clustering based on Dunn index

Similarly, we can extract the clustering that achieved the highest Dunn
score.

```{r best-clustering-dunn}
best_clustering_dunn <- hc_par[[optimal_indices["D"]]]
best_clustering_dunn
```

#### Retrieve the best clustering based on Calinski-Harabasz index

Finally, we can extract the clustering that achieved the highest
Calinski-Harabasz score.

```{r best-clustering-ch}
best_clustering_ch <- hc_par[[optimal_indices["CH"]]]
best_clustering_ch
```

This output helps you objectively compare different clustering outcomes
and select the most suitable solution for your data based on various
validity metrics.

## Performing Clustering with Optimal Configuration

Let's take a look at the clustering configuration that achieved the
maximum results according to the Calinski-Harabasz index.

```{r additional-hierarchical-clustering}
hc_3_ward <- tsclust(cryptos_list_wide,
  k = 3,
  type = "hierarchical",
  distance = "dtw",
  seed = 42,
  control = hierarchical_control(method = "ward.D2"),
  args = tsclust_args(dist = list(window.size = 7))
)

# View the clustering summary
hc_3_ward
```

```{r access-cluster-assignments-2}
# View cluster assignments for each time series
hc_3_ward@cluster
```

Plot the dendrogram for this clustering.

```{r plot-dendrogram}
# Plot the dendrogram
par(mar = c(0, 4, 2, 2))
plot(hc_3_ward,
  xlab = "", ylab = "", sub = "",
  main = "Hierarchical Clustering Dendrogram (DTW, Ward.D2)"
)
```

The line plots of the time series grouped by cluster show very distinct
patterns.

```{r fig-series-centroids-optimal}
#| fig-height: 3
#| fig-width: 8
#| fig-cap: "Time series clustered using DTW distance and Ward.D2 linkage."
# Plot time series and their centroids
plot(hc_3_ward, type = "sc") 
```

## Conclusion

`dtwclust` provides a modular and efficient framework for time-series
clustering in R, implementing various algorithms (especially DTW-related
ones) and allowing for customization and comparison. It serves as a
strong starting point for time-series clustering tasks.
