---
title: "Exploratory Data Analysis of Credit Risk Dataset"
author: "Aleksei Prishchepo"
date: "2025-09-15"
fig-format: svg
number-sections: true
format:
  html:
    toc: true
    toc-depth: 2
    toc-location: left
    highlight-style: tango
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(message = FALSE, warning = FALSE)
knitr::opts_chunk$set(dpi = 300, fig.width = 8)

library(tidyverse)
library(janitor)
library(skimr)
library(GGally)
library(ggplot2)
library(readr)
library(dplyr)
library(tidyr)
```

## Load Data

```{r load_data}
dataset <- readxl::read_excel("Credit_Risk_Dataset_Onyx_Data_September_25.xlsx")
```

## Basic Descriptives

```{r skim_data}
dataset |> skim()
```

```{r glimpse_data}
dataset |> glimpse()
```

Using latitude and longitude variables for predictions doesn't make much
sense, so we should remove them, along with `client_ID`, which serves
only as an identifier.

```{r}
dataset <- dataset |>
  select(-city_latitude, -city_longitude, -client_ID)
```

## Univariate Distributions

```{r }
#| fig-height: 6
#| fig-width: 6
# Numeric distributions
dataset |>
  select(where(is.numeric)) |>
  pivot_longer(everything()) |>
  ggplot(aes(x = value)) +
  facet_wrap(~name, scales = "free", ncol = 3) +
  geom_histogram(bins = 30, fill = "steelblue", color = "white") + 
  theme(axis.title = element_blank())
```

## Correlations

```{r fig-corrplot}
library(corrplot)

cor_data <- dataset |>
  select_if(is.numeric) |>
  cor(use = "pairwise.complete.obs")

cor_data |>
  corrplot(
    type = "upper", order = "hclust",
    tl.col = "black", tl.srt = 45
  )
```

There are three pairs of highly correlated variables. Let's check them
and drop redundant ones.

```{r}
cor(dataset$loan_percent_income, dataset$loan_to_income_ratio,
    use = "pairwise.complete.obs")
```

```{r}
cor(dataset$person_age, dataset$cb_person_cred_hist_length,
    use = "pairwise.complete.obs")
```

```{r}
cor(dataset$person_income, dataset$other_debt, use = "pairwise.complete.obs")
```

Obviously, `loan_percent_income` is in fact `loan_to_income_ratio`, so
let's drop it. Other two correlations are not perfect, so we keep them.

```{r}
dataset <- dataset |>
  select(-loan_percent_income)
```

## Missing Values

```{r fig-missing-upset}
#| fig-height: 3
#| fig-width: 4
library(naniar)

dataset |>
  select(person_emp_length, loan_int_rate) |>
  gg_miss_upset(
    sets.bar.color = "#56B4E9", point.size = 3,
    main.bar.color = "grey39", matrix.color = "grey39"
  )
```

There is no obvious pattern in missingness. We're going to impute
missing values using median.

```{r impute_missing}
dataset <- dataset |>
  mutate(across(where(is.numeric), ~ ifelse(is.na(.),
    median(., na.rm = TRUE), .
  )))
```

## Outliers

```{r fig-outliers}
#| fig-height: 6
#| fig-width: 6

dataset |>
  select(where(is.numeric)) |>
  pivot_longer(everything()) |>
  ggplot(aes(x = name, y = log(value))) +
  geom_boxplot(fill = "steelblue", 
               outlier.color = "red", outlier.size = 1) +
  theme(axis.title = element_blank(), 
        axis.text.x = element_text(angle = 45, hjust = 1))

```

There are outliers in `person_age` and `person_emp_length` that seem to
be dataset errors. There are also extreme values in `other_debt` and
`person_income`. We'll zoom in to see them better.

```{r fig-outliers-zoom}
require(gridExtra)
library(scales)

plot1 <- dataset |>
  select(where(is.numeric)) |>
  select(person_age, person_emp_length) |>
  pivot_longer(everything()) |>
  ggplot(aes(x = name, y = value)) +
  geom_boxplot(
    fill = "steelblue",
    outlier.color = "red", outlier.size = 1
  ) +
  theme(
    axis.title = element_blank(),
    axis.text.x = element_text(angle = 45, hjust = 1)
  )

plot2 <- dataset |>
  select(where(is.numeric)) |>
  select(other_debt, person_income) |>
  pivot_longer(everything()) |>
  ggplot(aes(x = name, y = log10(value))) +
  geom_boxplot(
    fill = "steelblue",
    outlier.color = "red", outlier.size = 1
  ) +
  scale_y_continuous(labels = trans_format(
    function(x) round(10^x, -1),
    label_number_auto()
  )) +
  theme(
    axis.title = element_blank(),
    axis.text.x = element_text(angle = 45, hjust = 1)
  )

grid.arrange(plot1, plot2, ncol = 2)

```

Let's correct `person_age` and `person_emp_length` by replacing outliers
with median. Other щutliers seem to be valid values. We're not going to
remove them.

```{r correct_outliers}
dataset <- dataset |>
  mutate(
    person_age = ifelse(person_age > 100,
      median(person_age, na.rm = TRUE),
      person_age
    ),
    person_emp_length = ifelse(person_emp_length > 70,
      median(person_emp_length, na.rm = TRUE),
      person_emp_length
    )
  )
```

Let's save the cleaned dataset for modelling.

```{r save_cleaned}
write_csv(dataset, "data_cleaned.csv")
```

## Selecting Predictors

We're going to build a logistic regression fit to predict loan default
(`loan_status` variable). `0` means no default, `1` means default.

We should exclude `loan_grade` variable as it is not independent. Also
`loan_int_rate` is likely to be set based on bank assumptions about
client, so we should exclude it as well.

```{r select_predictors}
dataset <- dataset |> select(-c("loan_grade", "loan_int_rate"))
```

Let's convert categorical variables to factors.

```{r convert_factors}
dataset <- dataset |>
  mutate(across(where(is.character), as.factor))
```

```{r linear-fit}
dataset <- dataset |> mutate(loan_status = (loan_status == 1))

fit <- glm(loan_status ~ .,
  family = binomial(link = "logit"),
  x = TRUE, data = dataset |> select(-city, -state)
)
summary(fit)
```

## Model Diagnostics

### Likelihood Ratio Test

```{r}
anova(fit, test="Chisq")
```

### Pseudo R-squared

```{r}
library(pscl)
pR2(fit)
```

```{r}
library(performance)
r2(fit)
```

### Overdispersion

Check for overdispersion:

```{r}
dispersion <- summary(fit)$deviance / summary(fit)$df.residual
dispersion
```

Dispersion is less than 1, so no overdispersion.

## Marginal Effects

```{r}
library(erer)

ma <- maBina(fit)$out

ma <- ma |>
  rownames_to_column("variable")
ma
```

## Updated Model with Encoded Categorical Variables

We need to deal with `city` and `country` variables as they cause
singularity in the model matrix. Let's encode them.

```{r encode-categorical}
one_hot_encoded <- model.matrix(~ city - 1, data = dataset)
data_encoded <- cbind(dataset, one_hot_encoded) 
one_hot_encoded <- model.matrix(~ country - 1, data = dataset)
data_encoded <- cbind(data_encoded, one_hot_encoded) |> select(-city, -country)
data_encoded |> glimpse()
```

```{r encoded-fit}
fit_encoded <- glm(loan_status ~ .,
  family = binomial(link = "logit"),
  x = TRUE, data = data_encoded
)
summary(fit_encoded)
```

## Encoded Model Diagnostics

### Likelihood Ratio Test

```{r}
anova(fit_encoded, test="Chisq")
```

### Pseudo R-squared

```{r}
library(pscl)
pR2(fit_encoded)
```

```{r}
library(performance)
r2(fit_encoded)
```

## AIC Optimization

Let's use stepwise selection to find an optimal set of predictors.

```{r stepwise-selection}

step_fit <- MASS::stepAIC(fit_encoded, direction = "both", trace = FALSE)
summary(step_fit)
step_fit$call
optimal_variables <- names(coef(step_fit))[-1]
optimal_variables
```

As soon as factor variables were encoded into multiple columns, we need
to make sure that all of them are included in the final model.

```{r}
final_variables <- c("loan_status")

for (var in optimal_variables) {
  if (var %in% (data_encoded |> colnames())) {
    final_variables <- c(final_variables, var)
  } else {
    cat(var, "is NOT in the dataset\n")
  }
}
```

Add back encoded categorical variables.

```{r}
final_variables |>
  c(final_variables, "person_home_ownership", "loan_intent", "cb_person_default_on_file")
```

```{r optimal-fit}
data_optimal <- data_encoded |> select(all_of(c("loan_status", final_variables)))
fit_optimal <- glm(loan_status ~ .,
  family = binomial(link = "logit"),
  x = TRUE, data = data_optimal
)
summary(fit_optimal)
```

### Pseudo R-squared

```{r}
pR2(fit_optimal)
```

```{r}
r2(fit_optimal)
```

## Selected Variables

```{r selected-variables}
selected_variables <- c(
  "loan_status", "person_home_ownership", "person_emp_length", "loan_intent", 
  "loan_amnt", "loan_to_income_ratio", "cb_person_default_on_file"
)

data_selected <- data_encoded |> select(all_of(selected_variables))

fit_selected <- glm(loan_status ~ .,
  family = binomial(link = "logit"),
  x = TRUE, data = data_selected
)
summary(fit_selected)

```

### Pseudo R-squared

```{r}
pR2(fit_optimal)
```

```{r}
r2(fit_optimal)
```

Let's compare all three models.

```{r model-comparison}

library(caret)   # for train, resampling
library(pROC)    # for AUC

set.seed(123)


data_encoded <- data_encoded |>
  mutate(loan_status = ifelse(loan_status == TRUE, "Yes", "No") |>
    as.factor())

data_optimal <- data_optimal |>
  mutate(loan_status = ifelse(loan_status == TRUE, "Yes", "No") |>
    as.factor())

data_selected <- data_selected |>
  mutate(loan_status = ifelse(loan_status == TRUE, "Yes", "No") |>
    as.factor())

colnames(data_encoded) <- make.names(colnames(data_encoded), unique = TRUE)
colnames(data_optimal) <- make.names(colnames(data_optimal), unique = TRUE)
colnames(data_selected) <- make.names(colnames(data_selected), unique = TRUE)


# Define cross-validation folds
ctrl <- trainControl(
  method = "cv",
  number = 5,              # 5-fold CV (use 10 for more stable estimates)
  classProbs = TRUE,       # needed for AUC
  summaryFunction = twoClassSummary
)

# Train full model
cv_full <- train(
  loan_status ~ ., 
  data = data_encoded,
  method = "glm",
  family = binomial,
  metric = "ROC",          # optimize on AUC
  trControl = ctrl
)

# Train stepAIC model (use the reduced dataset)
cv_step <- train(
  loan_status ~ ., 
  data = data_optimal,
  method = "glm",
  family = binomial,
  metric = "ROC",
  trControl = ctrl
)

# Train significant-only model
cv_sig <- train(
  loan_status ~ ., 
  data = data_selected,
  method = "glm",
  family = binomial,
  metric = "ROC",
  trControl = ctrl
)

# Compare CV AUC
cv_full
cv_step
cv_sig
```

Both the full model and the model using only significant variables show
very similar results. We'll use the model with selected variables as it
is more interpretable.

## Save Processed Data

```{r save-data}
write_csv(data_selected, "data_selected.csv")
```
